Velocity Map Imaging Spectroscopy of C$_2$H$^-$ and C$_2$D$^-$: a benchmark study of vibronic coupling interactions

### 2022-04-05 Aug2021-CH2-.tex
draft complete

\documentclass[journal=jpcafh,manuscript=article,layout=onecolumn, 12pt]{achemso}
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
\usepackage[T1]{fontenc}       % Use modern font encodings
\usepackage{amsmath}
\usepackage[dvipsnames]{xcolor}
% NB added command for in line cite
\newcommand{\onlinecite}[1]{\hspace{-1 ex} \nocite{#1}\citenum{#1}} 
%
\author{Benjamin~A.~Laws}
\email{b.laws@unsw.edu.au}
\affiliation{School of Chemistry, University of New South Wales, Sydney NSW 2052, Australia}
\alsoaffiliation{Research School of Physics, The Australian
	National University, Canberra ACT 2601, Australia}
\author{Zachariah~D.~Levey} 
\affiliation{School of Chemistry, University of New South Wales, Sydney NSW 2052, Australia}
\author{John F. Stanton}
\affiliation{Department of Chemistry, University of Florida, Gainesville, Florida 32611, United States}
\author{Timothy~W.~Schmidt} 
\affiliation{School of Chemistry, University of New South Wales, Sydney NSW 2052, Australia}
\author{Stephen~T.~Gibson}
\affiliation{Research School of Physics, The Australian
	National University, Canberra ACT 2601, Australia}
\title{Vibronic coupling in C$_2$H: implications for astrochemistry}
\abbreviations{PES,PAD,EA,eKE,FWHM,VMI}
\begin{document} 
\begin{abstract}

\end{abstract}
\section{Introduction}
Carbon monohydrides C$_{2n}$H, are a class of linear radicals that play an important role in combustion and interstellar chemistry~\cite{fre85,bou96,wil91,vui01}. These carbon chains have been observed in many interstellar environments, including planetary atmospheres~\cite{vui01,wil03,dob16}, comets~\cite{jac96}, dark clouds~\cite{ziu82,wil91,gup09}, and during heavy star formation~\cite{beu08,jia15}. Due to their relatively high abundance in a variety of astronomical conditions, they are believed to be promising candidates for Diffuse Interstellar Band (DIB) carriers~\cite{dou97,ful93,wat94,ful00,sch05,gup09}. The corresponding anions C$_{2n}$H$^-$ are also believed to play an important role in interstellar chemistry~\cite{mil17,gup09}. C$_6$H$^-$ was the first negative ion detected in space, after it was observed in two distinct interstellar molecular clouds IRC+10216 and TMC-1 in 2006~\cite{mcc06}. More recently, anions C$_4$H$^-$~\cite{cer07} and C$_8$H$^-$~\cite{bru07,rem07} have also been detected in a range of environments, include dark clouds~\cite{cor13}, prestellar cores,~\cite{sak10} and prostellar envelopes~\cite{sak07}.

Chain growth of the C$_{2n}$H radicals occurs predominately through acetylene addition,
\begin{equation}
\text{C}_{2n}\text{H} + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}+\text{H}_2.
\end{equation}
Conversely, the formation of anions in space is believed to be driven by radiative electron attachment, charge transfer, and dissociate electron attachment. Current astrochemical modelling of interstellar clouds, such as IRC+1026, typically underestimate the observed abundances of CH$_4^-$~\cite{mil17,cor13,her08}. Therefore, it has been proposed that ion-neutral reactions
 \begin{equation}
	\text{C}_{2n}\text{H}^- + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}^- + \text{H}_2.
\end{equation}
may be significant chain growth mechanisms that also needs to be included~\cite{mil17,bas19}. Modelling of the extraterrestrial planetary atmosphere of Saturn's moon Titan, suggests that both C$_2$H and C$_2$H$^-$ play an important role in the atmospheric chemistry~\cite{dob16,vui09,des17,vri18}. Multiple reaction pathways were included for the ion C$_2$H$^-$ including reactions with C$_2$H$_2$ and HCN.

Understanding interstellar chemistry relies critically on theory, to provide a link between astronomical observations and terrestrial laboratory studies. Microwave spectroscopy has successfully been used in this fashion to identify a large number of molecules in the interstellar medium (ISM). However for some species, particularly those with low abundances or no dipole moment, UV/vis spectroscopic methods are required for identification~\cite{mai97}. This creates a challenge for theory, as electronic spectra calculations are more sensitive to the level of theory employed than pure rotational spectra~\cite{for10}. This becomes even more challenging when vibronic coupling effects are introduced, as they can be exceedingly difficult to model.

@@

In this paper we study the vibronic spectrum of the ethynyl radical C$_2$H by employing High-Resolution Photoelectron Imaging (HR-PEI) in order to benchmark CFOUR \emph{ab-initio} vibronic-coupling calculations. C$_2$H is reported to be one of the most abundant molecules in the universe, and is the most thoroughly studied of the C$_{2n}$H species~\cite{wil91,hei99,for10}. Many different experimental techniques have been employed to try and resolve the complex vibronic spectrum, including electron spin resonance~\cite{coc64,jin85}, laser magnetic resonance~\cite{bro88,pfe96,sch98}, microwave and milli-meter wave spectroscopy~\cite{sas81,got83,end89,mul00}, infrared (matrix isolation~\cite{she87,jac87,for95} and Fourier Transform emission~\cite{ver88}) spectroscopy, laser induced fluorescence spectroscopy~\cite{hsu93,hsu95,chi99} and photoelectron spectroscopy~\cite{erv91,tay98,zho07}. C$_2$H has also received extensive theoretical attention to try and understand the experimental results, however the large number of possible vibronic levels means this has proven to be a challenge~\cite{per90,per91,per91b,per91c,per92,car00,tar03.tar04,for10,sta21}. In this work we demonstrate how construction of a quasiadiabatic Hamiltonian allows for the strength of vibronic interactions between coupled surfaces near a conical intersection to be calculated, in order to simulate electronic and vibronic spectra. By employing anion HR-PEI both the $^2\Sigma^+$ and $^2\Pi$ surfaces are mapped out on an equal footing from the anion $\tilde{X}^1\Sigma^+$ state, allowing for direct comparison to the \emph{ab-initio} modelling.

@@

To investigate the vibronic coupling interaction between these two nearby $\Sigma^+$ and $\Pi$ surfaces, deuterated ethynyl C$_2$D$^-$ was also studied. C$_2$D$^-$ ions were produced in a discharge of pure C$_2$D$_4$ gas and measured under the same experimental conditions as C$_2$H$^-$. An illustrative VMI of photodetachment at 355~nm from C$_2$D$^-$ is shown in Fig.~\ref{fig:1}(b). Similarly to Fig~\ref{fig:1}(a) two electronic transitions are observed. While the fast electrons appear similar in both images, a striking difference is observed in the slow electrons near the detector center. In Fig.~\ref{fig:1}(a) a series of weak rings are observed, however in Fig.~\ref{fig:1}(b) two dominant rings are seen.

The VMIs in Fig.~\ref{fig:1} were inverted using the Abel inversion methods detailed in pyAbel, to extract the corresponding photoelectron spectra presented in Fig.~\ref{fig:1}(c). Below $27,000~$cm$^{-1}$ on the $\tilde{X} ^2\Sigma^+$ surface the C$_2$H$^-$ and C$_2$D$^-$ spectra are very similar. Both are dominated by the origin transition, shifted by $\sim10~$cm$^{-1}$ in the deuterated spectrum, with progressions involving the v$_2$($\pi$) bending and v$_3$($\sigma$) CC stretch vibrational normal modes. This includes transitions involving an odd quanta of v$_2$ bending excitation ($2^{n+1}$) which should be totally forbidden within the harmonic oscillator approximation. However, the presence of these transitions in the spectra is likely an indicator of Herzberg-Teller (HT) vibronic coupling between the ground $^2\Sigma^+$ and nearby excited $^2\Pi$ electronic surfaces, as
\begin{equation}
\Sigma^+ \otimes \pi = \Pi. 
\end{equation}

@@

These interactions spread the electronic orgin of the $\tilde{A}\,^2\Pi$ state over several vibronic levels. Therefore, instead of assigning a defined origin, the observed peaks in the spectrum may be assigned to coupled admixtures of vibronic transitions involving the three potential energy surfaces,
\begin{equation}
\Psi_f = \sum_\xi \psi_e^\xi \sum_k C_{fk}^\xi\phi_{fkm}^\xi,
\label{eq:teller3} 
\end{equation}
where $\psi_e^\xi$ is the diabatic electronic wavefunction, and $\phi_{fkm}^\xi$ is the spin-rovibrational wavefunction. $\xi$ represents the electronic states used in the expansion ($\xi=\Sigma^+(1A'),~\Pi^+(2A'),~\Pi^-(1A'')$). A depiction of these three interacting surfaces is given in Fig.~\ref{fig:3}.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{figures/Fig3}
	\caption{Adiabatic potential energy surfaces of $\tilde{X}^2\Sigma^+$, $\tilde{A}^2\Pi^-$, and $\tilde{A}^2\Pi^+$ states, calculated using the method developed by Tarroni and Carter~\cite{tar03} (E$_{1A'}$, E$_{2A'}$, and E$_{1A''}$ in Eq.~\ref{eq:teller3}).}
	\label{fig:3}
\end{figure}

The adiabatic potential energy surfaces in Figure~\ref{fig:3} were calculated using the variational method of Tarroni and Carter. This has been used to calculate a multitude of admixed vibronic levels. In the literature, over 100 possible C$_2$H vibronic states have been calculated, up to 6,400~cm$^{-1}$ above the $\tilde{X} ^2\Sigma^+$ origin. However, assigning experimental spectra has remained a challenge, partly due to the large number of calculated levels providing multiple possible assignments for each experimentally observed transition.

@@

where the summations run over the dimensionless normal modes ($q_i$) that serve as the coordinate system for the problem.~\cite{sta21,car00} For convenience, the latter are chosen to be those of the anion, which considerably facilitates calculation of the spectral intensities.  In Eq.~(\ref{eq:calc1}), the diagonal terms of V (excluding those that carry the Renner-Teller coupling constant $\eta$) represent the quasidiabatic potential energy surfaces of the $\Sigma$, and the two components of the $\Pi$ state, chosen here as those in which the unpaired electron lies within or is perpendicular to an arbitrarily chosen plane, designated as A(a$'$) and A(a$''$), respectively. $\Delta_0^X$ is the separation of the anion and $^2\Sigma^+$ states at the origin of the coordinate system (the vertical electron detachment energy), and $\Delta_0^{A}$ is the gap between anion and $^2\Pi$ states at the same geometry.

For the modes of $\sigma$ symmetry ($q_1$ and $q_3$), the diabatic forces ($F_i$) and force constants ($F_{ij}$) coincide with those of the adiabatic potential energy surfaces.   However, for the bending mode, the diabatic (F$_{22}$) and adiabatic (f$_{22}$) force constants differ, and the parametrization is somewhat more involved.   For each component of the $\Pi$ state, either the 2a or 2b component of the bending vibration will maintain the A$'$ electronic symmetry that is needed to couple with the X state.  Designating these as 2a for the A(a$'$) state and 2b for the A(a$''$) state (as is implicit in Eq.~\ref{eq:calc1}), the diabatic force constants for the bending mode in the $^2\Sigma^+$ and $^2\Pi$ states can be written as

@@

When the potential in Eq.~(\ref{eq:calc1}) is diagonalized, the adiabatic states that are used for its parametrization are precisely recovered through terms second order in displacement. The linear diabatic coupling constants between the electronic states of interest were calculated analytically at the EOM-CCSDT/AN02 level of theory, using the CFOUR computational package and the quasidiabatic ansatz of Ichino~\emph{et al.}~\cite{ich09} Briefly, the EOM procedure is used to operationally define quasidiabatic states (those that relax according to a well-behaved reference state wavefunction) and the coupling constants are then evaluated as the first derivative of the off-diagonal elements of the electronic Hamiltonian in the basis defined by this representation. The diabatic force and coupling constants calculated to parametrize the KDC potential in Eq.~(\ref{eq:calc1}) are presented in Table~1.

@@

\subsection{Simulated Spectra}
The above method was employed to simulate the photoelectron spectrum of C$_2$H$^-$, with the calculated transitions shown in Figure~\ref{fig:C2H-plot}(a) alongside the experimental data at 355~nm. %Calculated transitions with $\Sigma$ symmetry are in blue, and transitions with $\Pi$ symmetry are in orange. 
On the $\tilde{X} ^2\Sigma^+$ surface, below 3,600~cm$^{-1}$, there is excellent agreement in both the transition positions and intensities between the simulated and experimental spectra. This includes the HT coupled (2$^{n+1}$) transitions with $\Pi$ symmetry, which would normally be missing from a simulation using standard \emph{ab-initio} approaches.

The positions of the calculated levels for the $\tilde{A}^2\Pi$ state have been shifted by 1,220~cm$^{-1}$ to account for the EOMIP-CCSDT calculations overestimating the effective Term energy. The experimental data shows that the electronic coupling interactions induce a splitting of the $\tilde{A} ^2\Pi$ state origin over 5 vibronic levels, spaced by $\sim 95~$cm$^{-1}$ (a-e). This splitting is also observed in the calculated spectrum, which correctly predicts 5 prominent transitions in this region. The relative intensities between the calculated transitions are also in good agreement with the experimental data. Peak d is slightly underestimated, however the relative intensity difference between peaks d and e may be partly explained by the Wigner threshold law, which will reduce the intensity of peak e in the experimental spectrum as it is close to the detachment threshold. The simulated spectrum also slightly underestimates the splitting between the vibronic levels, which may be linked to the overestimation of the gap between the $^2\Sigma$ and $^2\Pi$ surfaces.

\begin{figure}[th!]
	\includegraphics[width=0.6\textwidth]{figures/Fig5.pdf}
	\caption{cat}
	\label{fig:C2H-plot}
\end{figure}

The photoelectron spectrum of C$_2$D$^-$ was also examined using the same approach, with the calculated transitions shown in Figure~\ref{fig:C2H-plot}(b) alongside the 355~nm experimental data. Again, there is excellent agreement in the intensity and calculated positions of the transitions on the $\tilde{X} ^2\Sigma^+$ state surface, including the HT coupled levels. Near the $\tilde{A} ^2\Pi$ state origin, deuteration has a large impact on the experimental photoelectron spectrum. Instead of the 5 evenly split levels observed in C$_2$H, a single dominant peak (d) is now observed at 3,844~cm$^{-1}$ alongside a collection of weaker peaks (a-c) centred around 3,400~cm$^{-1}$. Based on previous spin-orbit splitting measurements, peak d is expected to have significant $\tilde{A} ^2\Pi$ character, with peaks a-c commonly assigned to vibronic levels with predominantly HT coupled $\tilde{X} ^2\Sigma^+$ character. This suggests that deuteration effectively dampens the coupling interaction between the surfaces.

Like the experimental spectra, a large change is observed in the simulated spectrum upon deuteration. The calculated intensity pattern of the vibronic levels around the $\tilde{A} ^2\Pi$ origin is a good match to the experimental data, with a single intense transition observed near peak d, and 3 prominent transitions predicted around the experimental peaks a-c. The calculated splitting between the levels is also similar, however the relative intensity of peak d appears to be slightly underestimated. %The results from Figure~\ref{fig:C2H-plot} confirm that the quasiadiabtic approach is able to accurately describe the vibronic interactions between the $\Sigma$ and $\Pi$ surfaces, including deuteration effects.

\begin{figure}[th!]
	\includegraphics[width=0.6\textwidth]{figures/266-plot.pdf}
	\caption{cat}
	\label{fig:266-plot}
\end{figure}
To investigate the effects of coupling on the higher excited vibronic levels on the$\tilde{A} ^2\Pi$ surface, the calculations were compared to the experimental photoelectron spectrum of C$_2$H$^-$ at 266~nm, as presented in Figure~\ref{fig:266-plot}. Gaussians were fitted to the transitions from Figure~\ref{fig:C2H-plot}(a) with a FWHM of 40~cm$^{-1}$ to match the resolution of the experimental data above threshold. Near 3,900~cm$^{-1}$ the electronic origin of the $\tilde{A} ^2\Pi$ state is split over 5 vibronic levels, as discussed above. A similar effect is also observed for the $3^1_0$ band, which becomes split over 3 vibronic levels labelled l, n, and o. The calculated spectrum correctly predicts the position and intensity of peaks n and o, but appears to strongly underestimate the intensity of peak l. Between the $0^0_0$ and $3^1_0$ bands a collection of weaker peaks are also observed (f,g,j,k). These are assigned to highly excited HT coupled $\tilde{X} ^2\Sigma^+$ transitions (j,k), and pure $\tilde{X} ^2\Sigma^+$ transitions (f,g). Consequently, peaks f and g will possess $\Sigma$ symmetry, and should have a different anisotropy to all of the other dominant transitions above 3,800~cm$^{-1}$. The position and intensity of these highly-excited coupled transitions are well described by the simulated spectrum.

The results from Figures~\ref{fig:C2H-plot} and~\ref{fig:266-plot} confirm that the quasiadiabtic approach is able to accurately describe the vibronic interactions between the $\Sigma$ and $\Pi$ surfaces, including deuteration effects. This validates the proposed vibronic interactions, and demonstrates how this method can be employed to decode and assign even complex spectra. Applying this approach to similar less-studied systems will produce reliable predictions for the position and intensity of dominant transitions, which will help guide the search for these molecules in laboratory experiments and astronomical observations.

%Overall, figures a-c show that the quasiadiabatic approach is a reliable method for simulating the spectra of molecules when strong vibornic coupling interactions are involved. This method may be expanded to other C$_n$H radicals that possess similar complexities, in order to help guide the search for new molecules to be found in space.

>>>>

\documentclass[journal=jpcafh,manuscript=article,layout=onecolumn, 12pt]{achemso}
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
\usepackage[T1]{fontenc}       % Use modern font encodings
\usepackage{amsmath}
\usepackage[dvipsnames]{xcolor}
% NB added command for in line cite
\newcommand{\onlinecite}[1]{\hspace{-1 ex} \nocite{#1}\citenum{#1}} 
%
\author{Benjamin~A.~Laws}
\email{b.laws@unsw.edu.au}
\affiliation{School of Chemistry, University of New South Wales, Sydney NSW 2052, Australia}
\alsoaffiliation{Research School of Physics, The Australian
	National University, Canberra ACT 2601, Australia}
\author{Zachariah~D.~Levey} 
\affiliation{School of Chemistry, University of New South Wales, Sydney NSW 2052, Australia}
\author{John F. Stanton}
\affiliation{Department of Chemistry, University of Florida, Gainesville, Florida 32611, United States}
\author{Timothy~W.~Schmidt} 
\affiliation{School of Chemistry, University of New South Wales, Sydney NSW 2052, Australia}
\author{Stephen~T.~Gibson}
\affiliation{Research School of Physics, The Australian
	National University, Canberra ACT 2601, Australia}
\title{Velocity Map Imaging Spectroscopy of C$_2$H$^-$ and C$_2$D$^-$: a benchmark study of vibronic coupling interactions}
\abbreviations{PES,PAD,EA,eKE,FWHM,VMI}
\begin{document} 
\begin{abstract}

\end{abstract}
High-resolution velocity map imaged photoelectron spectra of the ethynyl anions C$_2$H$^-$ and C$_2$D$^-$ are measured over a range of photodetachment wavelengths, to investigate the complex interactions between the close lying $\tilde{X}^2\Sigma^+$ and $\tilde{A}^2\Pi$ electronic states. Vibronic coupling calculations are performed by transforming to a quasidiabatic representation, parametrised by \emph{ab-initio} CFOUR calculations, to model the interplay between psuedo Jahn-Teller and Renner-Teller coupling. This approach is combined with electron anisotropy measurements to assign all of the observed vibronic structure, providing a framework that may be used to accurately simulate spectra of larger C$_{2n}$H monohydride carbon chains.

\section{Introduction}
Carbon monohydrides C$_{2n}$H, are a class of linear radicals that play an important role in combustion and interstellar chemistry~\cite{fre85,bou96,wil91,vui01}. These carbon chains have been observed in many interstellar environments, including planetary atmospheres~\cite{vui01,wil03,dob16}, comets~\cite{jac96}, dark clouds~\cite{ziu82,gup09}, and during heavy star formation~\cite{beu08,jia15}. Due to their relatively high abundance in a variety of astronomical conditions, they are believed to be promising candidates for Diffuse Interstellar Band (DIB) carriers~\cite{dou77,ful93,wat94,ful00,sch05,gup09}. The corresponding anions C$_{2n}$H$^-$ are also believed to play an important role in interstellar chemistry~\cite{mil17,gup09}. C$_6$H$^-$ was the first negative ion detected in space, after it was observed in two distinct interstellar molecular clouds IRC+10216 and TMC-1 in 2006~\cite{mcc06}. More recently, anions C$_4$H$^-$~\cite{cer07} and C$_8$H$^-$~\cite{bru07,rem07} have also been detected in a range of environments, include dark clouds~\cite{cor13}, prestellar cores,~\cite{sak10} and prostellar envelopes~\cite{sak07}.

Chain growth of the C$_{2n}$H radicals occurs predominately through acetylene addition,
\begin{equation}
\text{C}_{2n}\text{H} + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}+\text{H}_2.
\end{equation}
Conversely, the formation of anions in space is believed to be driven by radiative electron attachment, charge transfer, and dissociate electron attachment. Current astrochemical modelling of interstellar clouds, such as IRC+1026, typically underestimate the observed abundances of CH$_4^-$~\cite{mil17,cor13,her08}. Therefore, it has been proposed that ion-neutral reactions
 \begin{equation}
	\text{C}_{2n}\text{H}^- + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}^- + \text{H}_2.
\end{equation}
may be significant chain growth mechanisms that also needs to be included in astrochemical models~\cite{mil17,bas19}. Studies of the extraterrestrial planetary atmosphere of Saturn's moon Titan, suggest that both C$_2$H and C$_2$H$^-$ play an important role in the atmospheric chemistry~\cite{dob16,vui09,des17,vri18}. Multiple reaction pathways involving the C$_2$H$^-$ have been proposed, including reactions with C$_2$H$_2$ and HCN.

Our knowledge of interstellar chemistry relies critically on theory, to provide a link between astronomical observations and terrestrial laboratory studies. Microwave spectroscopy has successfully been used in this fashion to identify a large number of molecules in the interstellar medium (ISM). However for some species, particularly those with low abundances or no dipole moment, UV/vis spectroscopic methods are required for identification~\cite{mai97}. This creates a challenge for theory, as electronic spectra calculations are more sensitive to the level of theory employed than pure rotational spectra~\cite{for10}. This becomes even more challenging when vibronic coupling effects are introduced, as they can be exceedingly difficult to model.

@@

In this paper we study the vibronic spectrum of the ethynyl radical C$_2$H by employing High-Resolution Photoelectron Imaging (HR-PEI) in order to benchmark CFOUR~\cite{dev20}  \emph{ab-initio} vibronic-coupling calculations. C$_2$H is reported to be one of the most abundant molecules in the universe, and is the most thoroughly studied of the C$_{2n}$H species~\cite{wil91,hei99,for10}. Many different experimental techniques have been employed to try and resolve the complex vibronic spectrum, including electron spin resonance~\cite{coc64,jin85}, laser magnetic resonance~\cite{bro88,pfe96,sch98}, microwave and milli-meter wave spectroscopy~\cite{sas81,got83,end89,mul00}, infrared (matrix isolation~\cite{she87,jac87,for95} and Fourier Transform emission~\cite{ver88}) spectroscopy, laser induced fluorescence spectroscopy~\cite{hsu93,hsu95,chi99} and photoelectron spectroscopy~\cite{erv91,tay98,zho07}. C$_2$H has also received extensive theoretical attention to try and understand the experimental results, however the large number of possible vibronic levels means this has proven to be a challenge~\cite{per90,per91,per91b,per91c,per92,car00,tar03,tar04,for10,sta21}. In this work we demonstrate how construction of a quasidiabatic Hamiltonian allows for the strength of vibronic interactions between coupled surfaces near a conical intersection to be calculated, in order to simulate electronic and vibronic spectra. By employing anion HR-PEI both the $^2\Sigma^+$ and $^2\Pi$ surfaces are mapped out on an equal footing from the anion $\tilde{X}^1\Sigma^+$ state, allowing for direct comparison to the \emph{ab-initio} modelling.

@@

To investigate the vibronic coupling interaction between these two nearby $\Sigma^+$ and $\Pi$ surfaces, deuterated ethynyl C$_2$D$^-$ was also studied. C$_2$D$^-$ ions were produced in a discharge of pure C$_2$D$_4$ gas and measured under the same experimental conditions as C$_2$H$^-$. An illustrative VMI of photodetachment at 355~nm from C$_2$D$^-$ is shown in Fig.~\ref{fig:1}(b). Similarly to Fig~\ref{fig:1}(a) two electronic transitions are observed. While the fast electrons appear similar in both images, a striking difference is observed in the slow electrons near the detector center. In Fig.~\ref{fig:1}(a) a series of weak rings are observed, however in Fig.~\ref{fig:1}(b) the image is dominated by two distinct rings.

The VMIs in Fig.~\ref{fig:1} were inverted using the Abel inversion methods detailed in pyAbel~\cite{hic19}, to extract the corresponding photoelectron spectra presented in Fig.~\ref{fig:1}(c). Below $27,000~$cm$^{-1}$ on the $\tilde{X} ^2\Sigma^+$ surface the C$_2$H$^-$ and C$_2$D$^-$ spectra are a close match. Both are dominated by the origin transition, shifted by $\sim10~$cm$^{-1}$ in the deuterated spectrum, with progressions involving the v$_2$($\pi$) bending and v$_3$($\sigma$) CC stretch vibrational normal modes. This includes transitions involving an odd quanta of v$_2$ bending excitation ($2^{n+1}$) which should be totally forbidden within the harmonic oscillator approximation. However, the presence of these transitions in the spectra is likely an indicator of Herzberg-Teller (HT) vibronic coupling between the ground $^2\Sigma^+$ and nearby excited $^2\Pi$ electronic surfaces, as
\begin{equation}
\Sigma^+ \otimes \pi = \Pi. 
\end{equation}

@@

These interactions spread the electronic orgin of the $\tilde{A}\,^2\Pi$ state over several vibronic levels. Therefore, instead of assigning a defined origin, the observed peaks in the spectrum may be assigned to coupled admixtures of vibronic transitions involving the three potential energy surfaces,
\begin{equation}
\Psi_f = \sum_\xi \psi_e^\xi \sum_k C_{fk}^\xi\phi_{fkm}^\xi,
\label{eq:teller3} 
\end{equation}
where $\psi_e^\xi$ is the diabatic electronic wavefunction, and $\phi_{fkm}^\xi$ is the spin-rovibrational wavefunction. $\xi$ represents the electronic states used in the expansion ($\xi=\Sigma^+(1A'),~\Pi^+(2A'),~\Pi^-(1A'')$). A depiction of these three interacting surfaces is given in Fig.~\ref{fig:2}.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{figures/Fig3}
	\caption{Adiabatic potential energy surfaces of $\tilde{X}^2\Sigma^+$, $\tilde{A}^2\Pi^-$, and $\tilde{A}^2\Pi^+$ states, calculated using the method developed by Tarroni and Carter~\cite{tar03} (E$_{1A'}$, E$_{2A'}$, and E$_{1A''}$ in Eq.~\ref{eq:teller3}).}
	\label{fig:2}
\end{figure}

The adiabatic potential energy surfaces in Figure~\ref{fig:2} were calculated using the variational method of Tarroni and Carter~\cite{car00}. This has been used to calculate a multitude of admixed vibronic levels. In the literature, over 100 possible C$_2$H vibronic states have been calculated, up to 10,000~cm$^{-1}$ above the $\tilde{X} ^2\Sigma^+$ origin. However, assigning experimental spectra has remained a challenge, partly due to the large number of calculated levels providing multiple possible assignments for each experimentally observed transition.

@@

where the summations run over the dimensionless normal modes ($q_i$) that serve as the coordinate system for the problem.~\cite{sta21,car00} For convenience, the latter are chosen to be those of the anion, which considerably facilitates calculation of the spectral intensities.  In Eq.~(\ref{eq:calc1}), the diagonal terms of V (excluding those that carry the Renner-Teller coupling constant $\eta$) represent the quasidiabatic potential energy surfaces of the $^2\Sigma^+$, and the two components of the $^2\Pi$ state, chosen here as those in which the unpaired electron lies within or is perpendicular to an arbitrarily chosen plane, designated as A(a$'$) and A(a$''$), respectively. $\Delta_0^X$ is the separation of the anion and $^2\Sigma^+$ states at the origin of the coordinate system (the vertical electron detachment energy), and $\Delta_0^{A}$ is the gap between anion and $^2\Pi$ states at the same geometry.

For the modes of $\sigma$ symmetry ($q_1$ and $q_3$), the diabatic forces ($F_i$) and force constants ($F_{ij}$) coincide with those of the adiabatic potential energy surfaces.   However, for the bending mode, the diabatic (F$_{22}$) and adiabatic (f$_{22}$) force constants differ, and the parametrization is somewhat more involved.   For each component of the $^2\Pi$ state, either the 2a or 2b component of the bending vibration will maintain the A$'$ electronic symmetry that is needed to couple with the $^2\Sigma^+$ state.  Designating these as 2a for the A(a$'$) state and 2b for the A(a$''$) state (as is implicit in Eq.~\ref{eq:calc1}), the diabatic force constants for the bending mode in the $^2\Sigma^+$ and $^2\Pi$ states can be written as

@@

When the potential in Eq.~(\ref{eq:calc1}) is diagonalized, the adiabatic states that are used for its parametrization are precisely recovered through terms second order in displacement. The linear diabatic coupling constants between the electronic states of interest were calculated analytically at the EOM-CCSDT/AN02 level of theory, using the CFOUR computational package~\cite{dev20} and the quasidiabatic ansatz of Ichino~\emph{et al.}~\cite{ich09} Briefly, the EOM procedure is used to operationally define quasidiabatic states (those that relax according to a well-behaved reference state wavefunction) and the coupling constants are then evaluated as the first derivative of the off-diagonal elements of the electronic Hamiltonian in the basis defined by this representation. The diabatic force and coupling constants calculated to parametrize the KDC potential in Eq.~(\ref{eq:calc1}) are presented in Table~\ref{tab:1}.

@@

\begin{table}
	\caption{Parameters for the quasidiabatic Hamiltonian Eq.~(\ref{eq:calc1}) determined using CFOUR and the quasidiabatic ansatz.}
	\label{tab:1}
	\begin{tabular}{c |c c c c| c c c c }
		 & C$_2$H & &  & & C$_2$D & & & \\
		 parameter & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ \\
		 \hline\hline
		 Energy & 0 & 22431.21 &  26543.67 & 26543.67 & 0&	22431.21 & 26543.67	& 26543.67\\
		 f1 & 0	& -596.04 & 641.6 &	641.6 & 0& -596.04& 641.6 & 641.6\\
		 f2 & 0 & -1515.21 & 1257.07 & 1257.07 & 0 & -1401.53 & 1115.17 & 1115.17\\
		 f3 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
		 f4 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
         f11 & 3424.32 & 3487.03 & 3482.16 & 3482.16 & 2611.42 & 2636.57 & 2681.25 & 2681.25\\
         f12 & 0.00 & -28.32 & 31.46 & 31.46 & 0.00 & -48.38 & 55.40 & 55.40\\
         f22 & 1894.71 & 1830.83 & 2099.57 & 2042.21 & 1792.49 & 1747.62 & 1914.39 & 1914.39\\
         f33 & 607.09 & 87.57 & 1883.48 & 658.96 & 481.89 & 120.09 & 1503.88 & 473.55\\
         f44 & 607.09 & 87.57 & 658.96 & 1883.48 & 481.89 & 120.09 & 473.55 & 1503.88\\
         $\lambda$ & -1185.26 & & 770.77991 & -429.49481 & -1014.58 & & 620.701579 & -382.57 \\
         \end{tabular}
\end{table}

\subsection{Simulated Spectra}
The above method was employed to simulate the photoelectron spectrum of C$_2$H$^-$, with the calculated transitions shown in Figure~\ref{fig:3}(a) alongside the experimental data at 355~nm. %Calculated transitions with $\Sigma$ symmetry are in blue, and transitions with $\Pi$ symmetry are in orange. 
On the $\tilde{X} ^2\Sigma^+$ surface, below 3,600~cm$^{-1}$, there is excellent agreement in both the transition positions and intensities between the simulated and experimental spectra. This includes the HT coupled (2$^{n+1}$) transitions with $\Pi$ symmetry, which would normally be missing from a simulation using standard \emph{ab-initio} approaches.

The positions of the calculated levels for the $\tilde{A}^2\Pi$ state have been shifted by 1,220~cm$^{-1}$ to account for the EOMIP-CCSDT calculations overestimating the effective Term energy. The experimental data shows that the electronic coupling interactions induce a splitting of the $\tilde{A} ^2\Pi$ state origin over 5 vibronic levels, spaced by $\sim 95~$cm$^{-1}$ ($a-e$). This splitting is also observed in the calculated spectrum, which correctly predicts 5 prominent transitions in this region. The relative intensities between the calculated transitions are also in good agreement with the experimental data. Peak $d$ is slightly underestimated, however the relative intensity difference between peaks $d$ and $e$ may be partly explained by the Wigner threshold law~\cite{wig48}, which will reduce the intensity of peak e in the experimental spectrum as it is close to the detachment threshold. The simulated spectrum also slightly underestimates the splitting between the vibronic levels, which may be linked to the overestimation of the gap between the $^2\Sigma$ and $^2\Pi$ surfaces.

\begin{figure}[th!]
	\includegraphics[width=0.6\textwidth]{figures/Fig5.pdf}
	\caption{(a) Experimental photoelectron spectrum of C$_2$H$^-$ at 355~nm, compared to the simulated spectrum calculated using the quasidiabatic ansatz. Transitions with $\Sigma$ symmetry are shown in blue, transitions with $\Pi$ symmetry are shown in orange. (b) Experimental photoelectron spectrum of C$_2$D$^-$ at 355~nm, compared to the simulated spectrum. The labelling of peaks near the $\tilde{A}$ origin correspond to Table~S1.}
	\label{fig:3}
\end{figure}

The photoelectron spectrum of C$_2$D$^-$ was also examined using the same approach, with the calculated transitions shown in Figure~\ref{fig:3}(b) alongside the 355~nm experimental data. Again, there is excellent agreement in the intensity and calculated positions of the transitions on the $\tilde{X} ^2\Sigma^+$ state surface, including the HT coupled levels. Near the $\tilde{A} ^2\Pi$ state origin, deuteration has a large impact on the experimental photoelectron spectrum. Instead of the 5 evenly split levels observed in C$_2$H, a single dominant peak ($d$) is now observed at 3,844~cm$^{-1}$ alongside a collection of weaker peaks ($a-c$) centred around 3,400~cm$^{-1}$. Based on previous spin-orbit splitting measurements, peak d is expected to have significant $\tilde{A} ^2\Pi$ character, with peaks a-c commonly assigned to vibronic levels with predominantly HT coupled $\tilde{X} ^2\Sigma^+$ character~\cite{yan87,hsu95,chi99,wil11}. This suggests that deuteration effectively dampens the coupling interaction between the surfaces.
